---
title: "Course Project 1"
author: "Alberto Chong"
date: "12/29/2020"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### libraries  

We will attach the following libraries:  

```{r library, warning = FALSE, message = FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
```
  
  
### 1. Loading and preprocessing the data

#### Loading data
  
```{r loading}
activity <- read.csv(unzip(paste0(getwd(),"/activity.zip")))
```

#### transform data  

```{r transform}
activity <- activity %>% mutate(date = as.Date(date))
str(activity)
```

### 2. What is mean total number of steps taken per day?

#### Calculate the total number of steps taken per day

```{r, results='markup'}
Total <- activity %>% 
    group_by(date) %>% 
    summarise(total = sum(steps,na.rm = TRUE)) %>% 
    ungroup()
head(Total)
```

#### Make a histogram of the total number of steps taken each day

```{r}
hist(Total$total,
     main = "Histogram", 
     xlab = "steps per day")
```


### 3. Calculate and report the mean and median of the total number of steps taken per day
Mean of total number of steps per day
```{r}
mean(Total$total)
```

Median of total number of steps
```{r}
median(Total$total)
```

### 4. What is the average daily activity pattern?

#### Time series plot

```{r timeseries, warning=FALSE}
activity %>% group_by(interval) %>% 
  summarise(avg_steps = mean(steps,na.rm = T)) %>% 
  ggplot(aes(x = interval, y = avg_steps)) + 
  geom_line()
```

### 5. 5-minute interval that contains the maximum number of steps on average
```{r, warning=FALSE}
avg_interval <- activity %>% group_by(interval) %>% 
    summarise(avg_steps = mean(steps,na.rm = T))
avg_interval[which(avg_interval$avg_steps == max(avg_interval$avg_steps,na.rm = T)),]
```

### 6. Imputing missing values
#### Calculate and report the total number of missing values in the dataset
```{r}
sum(is.na(activity$steps))
```

#### Strategy for filling in all of the missing values in the dataset
```{r}
avg_day <- activity %>% 
    group_by(date) %>% 
    summarise(avg = round(mean(steps, na.rm = TRUE),0)) %>% 
    ungroup() %>% 
    replace_na(list(avg = 0))
head(avg_day)
```


#### Create a new dataset that is equal to the original dataset but with the missing data filled in
```{r}
imputed_activity <- activity %>% 
    inner_join(avg_day, by = "date") %>% 
    mutate(steps = ifelse(is.na(steps),avg,steps)) %>% 
    select(-avg)

head(imputed_activity)
```

### 7. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day

Histogram  
```{r}
imputed_Total <- imputed_activity %>% 
    group_by(date) %>% 
    summarise(total = sum(steps,na.rm = TRUE)) %>% 
    ungroup()
hist(imputed_Total$total,
     main = "Histogram (imputed data)", 
     xlab = "steps per day")
```

Mean: 
```{r}
mean(imputed_Total$total)
```
Median:  
```{r}
median(imputed_Total$total)
```

The values of mean an median didn't change from the estimates calculated from the first part of the assigment.  

### 8. Are there differences in activity patterns between weekdays and weekends?
#### Create a new factor variable in the dataset with two levels - "weekday" and "weekend"
```{r}
imputed_activity$weekday <- weekdays(imputed_activity$date)
imputed_activity <- imputed_activity %>% 
    mutate(class_weekday = ifelse(weekday %in% c("Saturday","Sunday"),"weekend","weekday"))
head(imputed_activity)
```
#### Make a panel plot containing a time series plot
```{r}
imputed_activity %>% 
  group_by(interval,class_weekday) %>% 
  summarise(avg_steps = mean(steps,na.rm = T)) %>% 
  ggplot(aes(x = interval, y = avg_steps)) + 
  geom_line() + 
  facet_wrap(~class_weekday, dir = "v")
```
